Genetic alterations and expression patterns of genes coding for FGF, FGFR, and FGFBP proteins in urothelial cancers

### 2024-12-05 report/markdown/report.Rmd
Update of figures, tables, analysis report

---
title: "FGF- and FGFR-coding gene alterations in urothelial cancers"
subtitle: "Short analysis report"
author: "Piotr Tymoszuk for the Department of Urology, Medical University of Innsbruck"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

@@

* Fibroblast growth factor receptor (FGFR) signaling has emerged as a promising therapy target in urothelial cancers with approval of the pan-FGFR inhibitor erdafitinib. However, effects of genetic alterations and modulation of expression of FGF- and FGFR-coding genes in urothelial carcinoma on tumor biology, therapy response, and clinical outcome are still not fully understood.

* The primary goal of the current project was to characterize frequency, co-occurrence, and clinical implications of genetic alterations in FGF- and FGFR-coding genes in large publicly available cohorts of urothelial cancers. The secondary goal was an attempt to establish subsets of urothelial cancer specimens based on differences in their profiles of somatic mutations and gene copy number variants.

* Four collectives of urothelial cancer patients were evaluated: the GENIE BLCA cohort, an urothelial carcinoma subset of the large GENIE project (n = `r nrow(genie$clinic)`) [@Sweeney2017], the predominantly muscle-invasive MSK IMPACT cohort (`r nrow(msk$clinic)`) [@Clinton2022], the muscle-invasive TCGA BLCA collective (n = `r nrow(tcga$clinic)`) [@Liu2018; @Robertson2017], and the IMvigor cohort of late stage, platinum-refractory cancers subjected to an anti-PD-L1 treatment (`r nrow(imvigor$clinic)`) [@Balar2017]. Detailed information on somatic mutations and their effects on protein sequence was provided for the GENIE BLCA, MSK IMPACT, and TCGA BLCA cohorts. Binarized indexes of presence/absence of somatic mutations in selected genes were available for the IMvigor cohort. Binarized data on gene amplifications and deletions were obtained for the GENIE BLCA, MSK IMPACT, and TCGA BLCA collectives. Gene expression quantification by RNA sequencing and assignment to consensus molecular classes [@Kamoun2020] were retrieved for the TCGA BLCA and IMvigor cohorts. Overall survival information was available for the MSK IMPACT, TCGA BLCA, and IMvigor cohorts; data on disease-specific and relapse-free survival was provided for the TCGA BLCA collective. Approximate overall survival was calculated for the GENIE BLCA project participants based on age at enrollment in years and follow-up time in days. Characteristic of the investigated collectives is presented in __Table \@ref(tab:tab-cohorts)__.

* Somatic mutations of _FGFR3_ belonged to the most common genetic features of urothelial cancers and were found in `r expl_genet$stats$mutation %>% filter(variable == 'FGFR3') %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of cancer samples. Somatic mutations of the remaining _FGFR1_, _FGFR2_, and _FGFR4_ receptor genes were less frequent (`r expl_genet$stats$mutation %>% filter(variable %in% c('FGFR1', 'FGFR2', 'FGFR4')) %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of cancer samples). Amplifications of the _FGF3_, _FGF4_, and _FGF19_ genes in urothelial cancers result from multiplication of the 11q13 chromosome region, which codes also for an essential oncogene _CCDN1_. Amplifications of _FGF3_, _FGF4_, and _FGF19_ were detected in `r expl_genet$stats$amplification %>% filter(variable %in% c('FGF3', 'FGF4', 'FGF19')) %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of specimens (__Figures \@ref(fig:fig-general-freq)__ and __\@ref(fig:fig-fgfr-freq)__, __Tables \@ref(tab:tab-genetics)__ and __\@ref(tab:tab-fgf-genetics)__).

* A great majority of somatic mutations of _FGFR1_, _FGFR2_, _FGFR3_, and _FGFR4_ was classified as missense mutations caused by single nucleotide polymorphisms. Particularly for _FGFR3_, they affected protein residues were frequently located apart from the Ig-like domains responsible for ligand biding and besides the kinase domains vital for activation of downstream signaling (__Figures \@ref(fig:fig-mutation-type)__ - __\@ref(fig:fig-protein-domain-split)__, __Table \@ref(tab:tab-fgfr-class)__). A more detailed analysis revealed the Ig-like domain 1 and kinase domain as the mutation hot-spots of _FGFR1_. The kinase domain was the prime target of mutations of _FGFR2_. Hinge regions between the Ig-like domains 1 and 2, as well as residues around the membrane-spanning region between the Ig-like domain 3 and the kinase domain were identified as mutational hot-spots of _FGFR3_. In case of _FGFR4_, the Ig-like domain 1, the hinge region between the Ig-like domain 3 and kinase domain, as well as the kinase domain were prime targets of mutations of _FGFR4_ (__Figures \@ref(fig:fig-fgfr1-position)__ - __\@ref(fig:fig-fgfr4-position)__). It can be hence speculated, that somatic mutations of FGF receptors may affect both the kinase activity and the conformation of the protein resulting in a higher affinity towards ligands or ligand-independent signaling.

* Co-occurrence and exclusivity of genetic features was investigated by analysis of pairwise Jacquard similarity coefficients for pairs of genetic alterations in urothelial cancer samples. Such analysis revealed little overlap between somatic mutations of _FGFR3_ and other common alterations such as _TP53_ and _RB1_ mutations, amplification of the 11q13 chromosome region hallmarked by copy number variants of _FGF3/4/19_ and _CCND1_, as well as deletion of the 9p21 chromosome region with the _MTAP_, _CDKN2A_, and _CDKN2A_ genes (__Figures \@ref(fig:fig-genetic-mds)__ - __Figure \@ref(fig:fig-fgfr-cont-hm)__).

* Concerning co-regulation of expression of FGF- and FGFR-coding genes, _FGFR1_ and _FGFR3_ mRNA levels were found to be weakly but significantly negatively associated. Increased expression of _FGFR1_ in the cancer tissue correlated significantly with upregulation of genes coding for multiple FGF ligands (_FGF1_, _FGF2_, _FGF5_, _FGF7_, _FGF9_, _FGF10_, and _FGF18_). This phenomenon was absent for _FGFR3_ (__Figure \@ref(fig:fig-fgfr-correlation)__). In a differential gene expression analysis of cancers with and without _FGFR3_ mutations, we observed that presence of _FGFR3_ mutations translated in significantly and strongly increased _FGFR3_ mRNA levels. By contrast, cancers with _FGFR3_ mutations had significantly lower expression of _FGF2_, _FGF5_, and _FGF7_ (__Figure \@ref(fig:fig-fgfr-dge)__, __Table \@ref(tab:tab-fgfr3-dge)__). This suggests two mechanisms leading to activation of pro-oncogenic FGFR signaling in urothelial carcinoma. In _FGFR3_ mutant tumors, FGFR signaling may be triggered by over-expression of FGFR3 protein, improved FGF-affinity, or even in an ligand-independent manner. Alternatively, concomitant over-expression of _FGFR1_ with its ligands may activated FGFR signaling in auto- or paracrine manner.

* No consistent differences in total mutation burdens (TMB), counts of protein-affecting somatic mutations, or counts of gene deletions were found for urothelial cancers with and without _FGFR3_ mutations. Gene amplification numbers were in turn significantly lower in _FGFR3_ mutated neoplasms as compared with _FGF3_ wild-type tumors. Effect size of this difference was small (__Figure \@ref(fig:fig-fgfr-tmb)__).

* There were no significant differences in age and gender distribution between patients with wild-type and _FGFR3_-mutated cancers. _FGFR3_ mutant cancers tended towards lower pathological stages, this effect was significant in the MSK IMPACT cohort but not in the TCGA BLCA collective (__Table \@ref(tab:tab-fgfr3-clinic)__). In the GENIE BLCA and MSK IMPACT cohorts, patients with _FGFR3_ mutated tumors demonstrated improved overall survival as compared with bearers of WT neoplasms. This phenomenon was also evident in the TCGA BLCA collective, it was only statistically significant upon exclusion of pT4 cancer patients (entire cohort: p = `r signif(fgfr_os$test$tcga$p_adjusted, 2)`, without pT4 cancers: p = `r signif(fgfr_os$test$tcga_wo_t4$p_adjusted, 2)`). There were no significant differences in disease-specific or relapse-free survival between the _FGFR3_ mutation strata in the TCGA BLCA cohort, which may however result from insufficient number of events of interest (tumor-related deaths and relapses, __Figures \@ref(fig:fig-fgfr-os)__ and __Figure \@ref(fig:fig-fgfr-rfs)__).

* Analysis of effects of amplification of the 11q13 chromosome region on expression of FGF- and FGFR-coding genes was possible solely for the TCGA BLCA collective. Presence of this genetic aberration resulting in amplification of the _FGF3_, _FGF4_, _FGF19_, and _CCND1_ genes translated into lower expression of _FGFR1_ and significantly higher levels of _FGF3_, _FGF4_, and _FGF19_ mRNA (__Figure \@ref(fig:fig-amp11q13)__). However, effects of those differences was weak and, in case of _FGF3_ and _FGF4_, the expression levels were extremely low as compared with other FGF ligand-coding genes. Hence, biological relevance of the 11q13 amplification for FGFR signaling remains unclear.

* Basal/squamous-like cancers followed by luminal papillary tumors constituted the largest consensus molecular classes [@Kamoun2020] of urothelial cancers in the TCGA BLCA and IMvigor cohorts (__Figure \@ref(fig:fig-fgfr-subtypes)A__). in both cohorts, _FGFR3_ mutated cancers were found to be significantly enriched in the papillary luminal class (TCGA: `r sub_genet$stats$tcga %>% filter(variable == 'FGFR3_mutation', consensusClass == 'LumP') %>% .$percent %>% signif(2)`%, IMvigor: `r sub_genet$stats$imvigor %>% filter(variable == 'FGFR3_mutation', consensusClass == 'LumP') %>% .$percent %>% signif(2)`% of specimens). In the TCGA BLCA cohort, amplifications of _FGF3_, _FGF4_, and _FGF19_ genes resulting from multiplication of the 11q13 chromosome region were found in luminal papillary (`r sub_genet$stats$tcga %>% filter(variable %in% c('FGF3_amplification', 'FGF4_amplification', 'FGF19_amplification'), consensusClass == 'LumP') %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`%), stroma-rich (`r sub_genet$stats$tcga %>% filter(variable %in% c('FGF3_amplification', 'FGF4_amplification', 'FGF19_amplification'), consensusClass == 'Stroma-rich') %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`%), and basal/squamous-like cancers (`r sub_genet$stats$tcga %>% filter(variable %in% c('FGF3_amplification', 'FGF4_amplification', 'FGF19_amplification'), consensusClass == 'Ba/Sq') %>% .$percent %>% .[1] %>% signif(2)`% of specimens, __Figures \@ref(fig:fig-fgfr-subtypes)__ and __\@ref(fig:fig-fgfr-subdetails)__, __Table \@ref(tab:tab-cons-genetics)__).

* mRNA levels of `r sub_dge$anova_significant %>% reduce(intersect) %>% map_chr(~paste0('_', .x, '_')) %>% collapse_and` were found to vary significant between the consensus molecular classes of urothelial carcinomas in both the TCGA BLCA and IMvigor data set (__Table \@ref(tab:tab-cons-dge)__). Expression of _FGFR1_ was the highest in stroma-rich and luminal non-specified tumors and the lowest in luminal papillary cancers. In turn, maximal mRNA levels of _FGFR3_ were detected in luminal papillary cancers and the lowest expression was observed in basal/squamous- and neuroendocrine-like tumors. Effect size of those differences was large (__Figure \@ref(fig:fig-sub-fgfr-expression)__). Expression of all significantly regulated _FGF_ ligands was the lowest in luminal papillary cancers. _FGF2_, _FGF7_, _FGF10_, and _FGF18_ transcript amounts followed the expression pattern of _FGFR1_ and were detected at the highest levels in stroma-rich and non-specified luminal cancers. _FGF5_ was expressed at the highest levels by basal/squamous-like, neuroendocrine-like and stroma-rich neoplasms. Effect sizes of differential regulation of these FGF-coding genes was large (__Figure \@ref(fig:fig-sub-fgf-expression1)__ and __\@ref(fig:fig-sub-fgf-expression2)__). Collectively, the differential gene expression analysis results support the model of activation of FGFR signaling propose above. While luminal papillary cancers may be driven by _FGFR3_ mutations and over-expression resulting in FGFR signaling with little dependence on the ligands, stroma-rich and non-specified luminal cancers may exploit auto- and paracrine signaling fueled by over-expression of _FGFR1_ and its ligands.

* By latent class analysis (LCA, called also latent class modeling), we could establish a classification scheme of urothelial cancers based on presence/absence of the most frequent genetic alterations detected in at least 5% of cancer specimens. This classification scheme with six 'genetic subsets' was developed in the GENIE BLCA training cohort and technically validated in the MSK IMPACT and TCGA BLCA cohort (__Figures \@ref(fig:fig-lca-devel)__ and __\@ref(fig:fig-lca-size)__). The 'mutRB1' subset (`r lca_eval$n_numbers %>% map(filter, clust_id == 'mutRB1') %>% map_dbl(~.x$percent) %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of specimens) was characterized by highly frequent _TP53_, _RB1_ and _ARID1A_ mutations, _E2F3_ amplifications, and nearly absent _FGFR3_ mutations, 11q13 amplification, and 9p21 deletions. The 'oligoMut' subset was hallmarked by frequent mutations of _TP53_ and the lowest overall mutation count. The key features of 'hyperMut' subset were the highest total mutation count, as well as frequent somatic mutations of _ARID1A_, _FGFR3_, _CREBBP_, _ERBB2_, _FBXW7_, and lacking copy number variants of 11q13 and 9p21. The 'del9p21' subset was characterized by relatively common deletions of the chromosome region 9p21, amplification of the 11q13 chromosome region, and highly frequent mutations of _TP53_ and _KDM6A_. In the 'mutFGFR3' subset, mutations of _FGFR3_ and _KDM6A_, and deletions of 9p21 were common, while mutations in _TP53_, _RB1_, and _ERBB2_, and gene amplifications were virtually absent. Finally, the 'ampMDM2' subset was characterized by the highest percentage of _MDM2_, highly common copy number variants of the 11p13 and 9p21 chromosome regions and essentially lacking mutations of _TP53_ and _RB1_ (__Figures \@ref(fig:fig-lca-genet1)__ - __\@ref(fig:fig-lca-details-cna2)__, __Table \@ref(tab:tab-lca-genetics)__).

* Numbers of mutations differed significantly between the genetic subsets with large effect size and were the highest in hyperMut cancers and the lowest in oligoMut and mutFGFR3 tumors. Gene deletion numbers were significantly higher in del9p21 and ampMDM2 than in the remaining genetic subsets in the GENIE BLCA and MSK IMPACT collectives (large effect size). Amplification counts were the highest in del9p21, ampMDM2, and mutRB1 subsets and the lowest in mutFGFR3 cancers; those differences were large and significant in all collectives (__Figures \@ref(fig:fig-lca-tmb)__ and __\@ref(fig:fig-lca-cna-count)__).

* We could observe significant differences in expression of _FGFR1_, _FGFR3_, and _FGFR3_, as well as _FGF2_, _FGF5_, FGF7_, and _FGF17_ genes between the genetic subsets of the TCGA BLCA cohort (__Table \@ref(tab:tab-lca-dge)__). Interestingly, mutFGFR3 cancers were found to over-express _FGFR3_ and, at the same time, to suppress _FGF2_, _FGF5_, and _FGF7_. The highest expression levels of _FGFR1_ were observed in mutRB1 and ampMDM2 genetic subsets; _FGFR2_ was found to be expressed in oligoMut, del9p21, and ampMDM2 at the highest levels (__Figures \@ref(fig:fig-lca-dge-fgfr)__ and __\@ref(fig:fig-lca-dge-fgf)__).

>>>>

---
title: "Genetic alterations and expression patterns of genes coding for FGF, FGFR, and FGFBP proteins in urothelial cancers"
subtitle: "Short analysis report"
author: "Piotr Tymoszuk for the Department of Urology, Medical University of Innsbruck"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

@@

* Fibroblast growth factor receptor (FGFR) signaling has emerged as a promising cancer therapy target with approval of the pan-FGFR inhibitor erdafitinib for treatment of urothelial carcinoma [@Korc2009]. However, effects of genetic alterations and modulation of expression of FGF- and FGFR-coding genes in urothelial carcinoma on tumor biology, therapy response, and clinical outcome are still not fully understood. In particular, it is still not completely clear how and to which extent development and aggressiveness is driven by FGFR signaling in various form of urothelial carcinoma reflected by e.g. consensus molecular subtypes [@Kamoun2020].

* The primary goal of the current project was to characterize frequency, co-occurrence, and clinical implications of genetic alterations and differential expression of FGF-, FGFR-, and FGF binding protein (FGFBP) coding genes (__Table \@ref(tab:tab-genes)__) [@Korc2009; @Turati2022; @Bennett1995; @Gubbiotti2016; @Mundhenke2002; @Midorikawa2003; @Leali2012; @Xie2020; @Hagihara2000; @Pan2021; @Mohammadi2005; @Hu2009a] in large publicly available cohorts of urothelial cancers. The secondary goal was an attempt to establish subsets of urothelial cancer specimens based on differences in their profiles of somatic mutations and gene copy number variants.

* Five collectives of urothelial cancer patients were evaluated: the GENIE BLCA cohort, an urothelial carcinoma subset of the large GENIE project (n = `r nrow(genie$clinic)`) [@Sweeney2017], the predominantly muscle-invasive MSK IMPACT cohort (`r nrow(msk$clinic)`) [@Clinton2022], the muscle-invasive TCGA BLCA collective (n = `r nrow(tcga$clinic)`) [@Liu2018; @Robertson2017], the IMvigor cohort of late stage, platinum-refractory cancers subjected to an anti-PD-L1 treatment (`r nrow(imvigor$clinic)`) [@Balar2017], and the BCAN cohort of metastatic urothelial cancers [@Damrauer2022]. Detailed information on somatic mutations and their effects on protein sequence was provided for the GENIE BLCA, MSK IMPACT, TCGA BLCA, and BLCA cohorts. Binarized indexes of presence/absence of somatic mutations in selected genes were available for the IMvigor cohort. Binarized data on gene amplifications and deletions were obtained for the GENIE BLCA, MSK IMPACT, and TCGA BLCA collectives. Gene expression quantification by RNA sequencing and assignment to consensus molecular classes [@Kamoun2020] were retrieved for the TCGA BLCA, IMvigor, and BCAN cohorts. Overall survival information was available for the MSK IMPACT, TCGA BLCA, IMvigor, and BCAN cohorts; data on disease-specific and relapse-free survival was provided for the TCGA BLCA collective. Approximate overall survival was calculated for the GENIE BLCA project participants based on age at enrollment in years and follow-up time in days. Characteristic of the investigated collectives is presented in __Table \@ref(tab:tab-cohorts)__.

* Somatic mutations of _FGFR3_ belonged to the most common genetic features of urothelial cancers and were found in `r expl_genet$stats$mutation %>% filter(variable == 'FGFR3') %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of cancer samples. Somatic mutations of the remaining _FGFR1_, _FGFR2_, and _FGFR4_ receptor genes were less frequent (`r expl_genet$stats$mutation %>% filter(variable %in% c('FGFR1', 'FGFR2', 'FGFR4')) %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of cancer samples). Amplifications of the _FGF3_, _FGF4_, and _FGF19_ genes in urothelial cancers result from multiplication of the _11q13_ chromosome region, which codes also for an essential oncogene _CCDN1_. Amplifications of _FGF3_, _FGF4_, and _FGF19_ were detected in `r expl_genet$stats$amplification %>% filter(variable %in% c('FGF3', 'FGF4', 'FGF19')) %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of specimens (__Figure \@ref(fig:fig-general-freq)__ and __\@ref(fig:fig-fgfr-freq)__, __Table \@ref(tab:tab-genetics)__ and __\@ref(tab:tab-fgf-genetics)__).

* Mutations of _FGFR1_, _FGFR2_, _FGFR3_, and _FGFR4_ were predominantly classified as missense mutations caused by single nucleotide polymorphisms (__Figure \@ref(fig:fig-mutation-type)__ and __\@ref(fig:fig-variant-type)__, __Table \@ref(tab:tab-fgfr-class)__).  For _FGFR1_, the mutation hot-spots were located in the Ig-like domain 1 and kinase domain. The kinase domain was the prime target of mutations of _FGFR2_. For _FGFR3_, they affected protein residues were located between Ig-like domains 1 and 2 responsible for ligand biding, or at the membrane-spanning hinge between the ligand-binding region and the kinase domain responsible for activation of the downstream signaling. In case of _FGFR4_, the Ig-like domain 1, the hinge region between the Ig-like domain 3 and kinase domain, as well as the kinase domain were prime targets of mutations of _FGFR4_ (__Figures \@ref(fig:fig-protein-domain-split)
__ - __\@ref(fig:fig-fgfr4-position)__, __Table \@ref(tab:tab-fgfr-class)__). It can be hence speculated, that somatic mutations of the _FGFR_ genes may affect both the kinase activity and the conformation of the protein resulting in a higher affinity towards ligands or ligand-independent signaling.

* Co-occurrence and exclusivity of genetic features was investigated by analysis of pairwise Jacquard similarity coefficients for pairs of genetic alterations in urothelial cancer samples. Such analysis revealed little overlap between somatic mutations of _FGFR3_ and other common alterations such as _TP53_ and _RB1_ mutations, amplification of the 11q13 chromosome region hallmarked by copy number variants of _FGF3/4/19_ and _CCND1_, as well as deletion of the 9p21 chromosome region with the _MTAP_, _CDKN2A_, and _CDKN2A_ genes (__Figures \@ref(fig:fig-genetic-mds)
__ - __Figure \@ref(fig:fig-fgfr-cont-hm)__).

* Co-expression of FGFR-, FGF-, and FGFBP-coding genes was explored by correlation and network analyses, which revealed a high degree co-regulation of the ligands, receptors, and binding proteins in the urothelial cancer tissue (__Figure \@ref(fig:fig-fgfr-correlation)__ and __Figure \@ref(fig:fig-fgfr-networks)__). Especially the network analyses point towards strong co-regulation of expression of _FGFR1_ with genes of its ligands (_FGF2_, _FGF7_, _FGF10_), and co-receptors such as _SDC2_, _DCN_ or _HSPG2_. Importantly, _FGFR1_ (hub score: `r expl_exnet$stats %>% map(filter, name == 'FGFR1') %>% map_dbl(~.x$hub_score) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, general correlation strength with maximum of 1, degree: `r expl_exnet$stats %>% map(filter, name == 'FGFR1') %>% map_dbl(~.x$degree) %>% range %>% signif(2) %>% paste(collapse = ' to ')`), _FGF2_ (hub score: `r expl_exnet$stats %>% map(filter, name == 'FGF2') %>% map_dbl(~.x$hub_score) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, degree: `r expl_exnet$stats %>% map(filter, name == 'FGF2') %>% map_dbl(~.x$degree) %>% range %>% signif(2) %>% min`), _FGF7_ (hub score: `r expl_exnet$stats %>% map(filter, name == 'FGF7') %>% map_dbl(~.x$hub_score) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, degree: `r expl_exnet$stats %>% map(filter, name == 'FGF7') %>% map_dbl(~.x$degree) %>% range %>% signif(2) %>% paste(collapse = ' to ')`), and _FGF10_ (hub score: `r expl_exnet$stats %>% map(filter, name == 'FGF10') %>% map_dbl(~.x$hub_score) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, degree: `r expl_exnet$stats %>% map(filter, name == 'FGF10') %>% map_dbl(~.x$degree) %>% range %>% signif(2) %>% paste(collapse = ' to ')`) were identified as candidate highly connected hubs of the co-expression networks, which suggests that FGFR1 signaling as a critical player in urothelial cancer biology. In turn, _FGFR2_ was found to be weakly co-regulated with the non-canonical FGF receptor gene, _TGFBR3_. _FGFR3_ levels correlated weakly with levels of _SDC1_ and _GPC1_ transcripts that code for proteoglycan FGF binding proteins.

* In a differential gene expression analysis of cancers with and without _FGFR3_ mutations, presence of _FGFR3_ mutations translated to significantly and strongly increased _FGFR3_ mRNA levels, and, to a lesser extent, upregulation of the syndecan gene _SDC1_. By contrast, cancers with _FGFR3_ mutations had significantly lower expression of _FGFR1_, _FGF2_, _FGF5_, and _FGF7_ (__Figure \@ref(fig:fig-fgfr-dge-volcano)__ and __\@ref(fig:fig-fgfr-dge-box)__, __Table \@ref(tab:tab-fgfr3-dge)__). No consistent differences in counts of protein-affecting somatic mutations, or counts of gene deletions were found for urothelial cancers with and without _FGFR3_ mutations. Gene amplification numbers were in turn significantly lower in _FGFR3_ mutated neoplasms as compared with _FGF3_ wild-type tumors. Effect size of this difference was moderate (__Figure \@ref(fig:fig-fgfr-tmb)__).

* There were no significant differences in age and gender distribution between patients with wild-type and _FGFR3_-mutated cancers. _FGFR3_ mutant cancers tended towards lower pathological stages, this effect was significant in the MSK IMPACT cohort but not in the TCGA BLCA collective (__Table \@ref(tab:tab-fgfr3-clinic)__). In the GENIE BLCA and MSK IMPACT cohorts, patients with _FGFR3_ mutated tumors demonstrated improved overall survival as compared with bearers of WT neoplasms. This phenomenon was also evident in the TCGA BLCA collective and was statistically significant upon exclusion of pT4 cancer patients (entire cohort: p = `r signif(fgfr_os$test$tcga$p_adjusted, 2)`, without pT4 cancers: p = `r signif(fgfr_os$test$tcga_wo_t4$p_adjusted, 3)`). There were no significant differences in disease-specific or relapse-free survival between the _FGFR3_ mutation strata in the TCGA BLCA and BCAN cohorts, which may, however, result from insufficient number of relapses and tumor-related deaths (__Figures \@ref(fig:fig-fgfr-os)__ and __Figure \@ref(fig:fig-fgfr-rfs)__).

* Amplification of the _11q13_ chromosome region was one of the most frequent genetic alterations in urothelial cancers and led to multiplication of _FGF3_, _FGF4_, and _FGF19_ genes. Analysis of effects of this amplification expression of FGFR-, FGF- and FGFBP-coding genes was possible solely for the TCGA BLCA collective. Cancer samples with the _11q13_ amplification had significantly higher levels of _FGF3_, _FGF4_, and _FGF19_ mRNA in the cancer tissue. _FGFR1_ was downregulated without formal statistical significance (__Figure \@ref(fig:fig-amp11q13)__). Effect sizes of those differences were weak or moderate, and, in case of _FGF3_ and _FGF4_, the expression levels were extremely low as compared with other FGF-coding genes. Hence, biological relevance of the _11q13_ amplification for FGFR signaling remains unclear.

* Basal/squamous-like cancers followed by luminal papillary and stroma-rich tumors constituted the largest consensus molecular classes of urothelial cancers [@Kamoun2020] in the TCGA BLCA, IMvigor, and BCAN cohorts. In all three collectives, _FGFR3_ mutations were found to be significantly enriched in luminal papillary cancer (TCGA: `r sub_genet$stats$tcga %>% filter(variable == 'FGFR3_mutation', consensusClass == 'LumP') %>% .$percent %>% signif(2)`%, IMvigor: `r sub_genet$stats$imvigor %>% filter(variable == 'FGFR3_mutation', consensusClass == 'LumP') %>% .$percent %>% signif(2)`%, BCAN: `r sub_genet$stats$bcan %>% filter(variable == 'FGFR3_mutation', consensusClass == 'LumP') %>% .$percent %>% signif(2)` of specimens). In the TCGA BLCA cohort, amplifications of _FGF3_, _FGF4_, and _FGF19_ genes resulting from multiplication of the _11q13_ chromosome region were found in luminal papillary (`r sub_genet$stats$tcga %>% filter(variable %in% c('FGF3_amplification', 'FGF4_amplification', 'FGF19_amplification'), consensusClass == 'LumP') %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`%), stroma-rich (`r sub_genet$stats$tcga %>% filter(variable %in% c('FGF3_amplification', 'FGF4_amplification', 'FGF19_amplification'), consensusClass == 'Stroma-rich') %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`%), and basal/squamous-like cancers (`r sub_genet$stats$tcga %>% filter(variable %in% c('FGF3_amplification', 'FGF4_amplification', 'FGF19_amplification'), consensusClass == 'Ba/Sq') %>% .$percent %>% .[1] %>% signif(2)`% of specimens) (__Figure \@ref(fig:fig-fgfr-subtypes)__ and __\@ref(fig:fig-fgfr-subdetails)__, __Table \@ref(tab:tab-cons-genetics)__).

* Analysis of differential expression of FGFR-, FGF-, and FGFBP-coding genes was conducted for the major consensus molecular subtypes of the TCGA BLCA, IMvigor, and BCAN cohort. Transcript levels of n = `r length(unique(unlist(sub_dge$common_significant)))` out of `r length(reduce(sub_dge$variables, union))` investigated genes (__Table \@ref(tab:tab-cons-dge)__). Among them, _FGFR3_, _SDC1_, _TGFBR3_, _GPC3_, and _SDC3_ were upregulated specifically in luminal papillary cancers. Overexpression of _KL_, a gene coding for [Klotho](https://www.genecards.org/cgi-bin/carddisp.pl?gene=KL) FGF-binding protein, was characteristic for luminal unstable tumors. Genes coding for FGFR protein, its ligands (_FGF7_, _FGF14_, _FGF10_, _FGF2_, _FGF9_, _FGF18_), and interaction partners (_DCN_, _HSPG2_, _SDC2_, _PTX3_) peaked in the stroma-rich class. High expression of _FGF5_, _FGF11_, and genes coding for binding proteins and FGF co-receptors, _CD44_, _FGFBP1_, _GPC1_ and _TNFAIP6_, was a hallmark of basal/squamous-like cancers (__Figures \@ref(fig:fig-sub-fgfr-expression)__ - __Figure \@ref(fig:fig-sub-fgf-expression2)__).

* Encouraged by the profound and widespread differences in expression of FGFR-, FGF-, and FGFBP-coding genes, we asked if and to which extent expression of those genes can explain or emulate the consensus molecular subsets proposed by Kamoun and colleagues [@Kamoun2020]. To answer this intricate question, we resorted to machine learning with two unrelated algorithms, Elastic Net regularized linear regression [@Friedman2010; @Zou2005] and Random Forest tree ensemble [@Wright2017; @Breiman2001]. The models were trained in a synthetic data set of ComBat-processed $log_2$-transformed gene expression derived from the TCGA BLCA cohort. To improve identification of the minority molecular classes, LumU and stroma-rich cancers, their observations were amplified (over-sampled) in the training data set with the SMOTE algorithm [@Torgo2014; @Chawla2002; @Torgo2016]. Performance and calibration of the machine learning models was evaluated by comparison of the model-predicted with the actual consensus molecular classes computed by a classification tool of _consensusMIBC_ R package (ground truth) [@Kamoun2020] in the unmodified TCBA BLCA data set and the independent validation IMvigor and BCAN cohorts.

* As presented in __Figure \@ref(fig:fig-sub-elnet)A__ and __Table \@ref(tab:tab-ml-eval)__, the machine learning models predicted the molecular class assignment with very high accuracy, independently of the algorithm. Notably, fractions of correctly predicted cancer samples were lower for the luminal unstable class as compared with the remaining molecular classes (__Figure \@ref(fig:fig-sub-elnet)B__). Slightly worse reliability of the models at predicting luminal unstable and stroma-rich samples in the validation cohorts was also evident in results of the calibration analyses (__Figure \@ref(fig:fig-sub-cal-elnet)__ and __\@ref(fig:fig-sub-cal-ranger)__). __Collectively, the machine learning data suggests that the consensus molecular classification by Kamoun et al. can be reproduced by differences in expression of `r length(ml_globals$variables)` genes coding for FGFR, FGF, and FGFBP proteins.__ Notably, the classification tool developed by Kamoun and colleagues and distributed via _consensusMIBC_ R package employs `r consensusMIBC::centroids %>% nrow` gene variables. Among them, only `r intersect(consensusMIBC::centroids$hgnc_symbol, ml_globals$variables) %>% map_chr(~paste0('_', .x, '_')) %>% collapse_and` were shared with our machine learning models. The most influential genes for prediction of consensus molecular subtypes in our machine learning approach are presented in __Figure \@ref(fig:fig-sub-imp)__, and __Tables \@ref(tab:tab-ml-imp-elnet)__ - __\@ref(tab:tab-ml-imp-ranger)__.

* Furthermore, we asked about prognostic significance of differences in FGFR, FGF, and FGFBP gene expression. To answer it, we constructed a multi-parameter Elastic Net Cox model [@Zou2005; @Friedman2010] in the TCGA BLCA cohort (__Table \@ref(tab:tab-surv-eval)__ and __\@ref(tab:tab-surv-imp)__). ComBat-adjusted $log_2$-transformed expression levels were fed into the model as first- and second-order terms. Among them, `r rep_tabs$surv_imp %>% filter(.data[['Coefficient beta']] != 0, !duplicated(.data[['Gene symbol']])) %>% nrow` genes had non-zero coefficients, i.e. contributed substantially to prediction of overall survival (__Figure \@ref(fig:fig-surv-ml)A__). The model delivered reliable predictions in the TCGA BLCA training collective and in the IMvigor validation cohort, but not in the BCAN data set (__Figure \@ref(fig:fig-surv-ml)BC__). In an univariable, model-free screening of the genes of interest as potential prognostic markers, we found reproducibly shorter overall survival of patients with high expression of _TNFAIP6_, _GPC1_, _FIBP_, _FGF11_, and _FGFBP1_ in the tumor tissue as compared with low expressors (__Figure \@ref(fig:fig-surv-km1)__ and __\@ref(fig:fig-surv-km2)__, __Table \@ref(tab:tab-surv-km)__). Taken together, the FGFR, FGF, and FGFBP transcriptome bears some potential for prediction of overall survival in urothelial cancer, but the models and standalone markers need additional validation, especially with inclusion of confounders such as age, gender, and staging.
